Cheminformatics: The Computer Science of Chemical Discovery

### 2011-11-18 rebuttal.tex

\documentclass[11pt,a4paper]{article}
\usepackage{times}
\usepackage{fullpage}
\setlength{\parindent}{0cm}

\begin{document}
\newcommand{\comment}[1]{\textit{``\ldots #1''}\par\vspace{0.5em}}
\newcommand{\response}[1]{#1\vspace{1em}}

We would like to thank the anonymous reviewers for their constructive
comments which have served to significantly improve our manuscript. 
We have updated the main text in several places and have
also included a supplemental document that provides a brief
description of terminology and a list of some historic milestones in
the field. The following sections provide a detailed response to
individual reviewer comments.
\vspace{2em}

@@

\comment{The authors mention limitations of SMILES implementations,
  but do not mention canonicalization in this context, often using the
  Morgan algorithm published in 1965 in J. Chem. Doc.}
\response{We have updated the text on page 3 to note the Morgan
  algorithm in the context of chemical structure representation}

\comment{The authors state that 3D information is lost when only
  considering the molecular graph. However, they do not make it clear
  that the 3D space is implicitly encoded in the molecular graph –
  this is essential to cover properly.}
\response{We have update the text on page 2 to note that 3D
  information is implicit in 2D representations}

\comment{ I entirely dispute that cheminformatics methods have been
  'closely guarded secrets of companies...' since many methods have been
  published in the public domain. The authors suggest that only in the
  last decade has the field gained access to freely available
  software, etc.}
\response{We agree that secret was probably not the best choice of
  words. However we do believe that much cheminformatics software and
  data was proprietary till recently. We have updated the text on
  pages 7 and 8 to expand on this and provide a more detailed
  discussion}

\comment{If this manuscript were to be published it would be
  completely rewriting the history of cheminformatics}
\response{While we agree with the reviewer that some key milestones in
  the history of cheminformatics were not included in the original
  article, we believe that the addition of the supplemental history
  addresses this problem. Furthermore, our original aim was \emph{not}
  to provide a complete historical overview of the field. Others have
  already provided such overviews. Our aim was to present a brief
  overview of \emph{current topics} in cheminformatics, selected because we
  believed that they represent on going challenges for the field and
  thus amenable to coss-disciplinary efforts.}

@@

%%%%%%%%%%%%%%%%%%%%%%%

\fbox{\textbf{Reviewer 3}}

\comment{The manuscript provides a very specific overview on the
  aspects of the open source community in cheminformatics. }
\response{As we have noted below, the current article is not a
  comprehensive historical review of the field. We have included a
  supplement  that does provide a very brief overview of historic
  milestones. Furthermore, we have updated the text to note commercial
  vendors and solutions; but we still stress on Open Source and Open
  Access solutions to allow reasders to explore the problems and
  topics we have discussed with a minimum of hindrances}

@@

\comment{The manuscript is poor in relevant citations with none of the
  classical references provided, such as Morgan 1965, Weininger 1988,
  Johnson and Maggiora 1990, Barnard 1993, Willett et al. 1998, etc.}
\response{The text has been updated to include the ``classic''
  references. In addition the supplemental history document lists a
  number of milestones in cheminformatics and their relevant references.}

\comment{Molecular docking is also not mentioned, but is regarded as
  one of the key methods applied in cheminformatics. Docking should be
  included, along with other molecular modelling methods such as
  pharmacophores, shape searching, molecular dynamics, as a distinct
  section.}
\response{These methods are mentioned on page 5 (column
  1). However, we have chosen not to place a more in depth discussion
  of these methods in their own section in the main document for a few
  reasons. First, due to space limitations this would require us to
  compress or drop other sections which we believe to be more
  relevant. Second, we are of the view that docking, for example, is
  not really a cheminformatics technique, though it makes extensive
  use of cheminformatic methods. Instead we place it under molecular
  modeling. We realize that this is a somewhat subjective
  ``classification'' of methods. However we do make explicit note of
  these methods in the supplemental document where we make a
  distinction between cheminformatics and related subject areas such
  as quantum mechanics and molecular modeling.}

@@

\comment{Proteochemometric methods are mentioned a couple of times but
  these are largely un-noticed by the community currently. I do not
  think that it is necessarily appropriate to include these as
  tried-and-tested methods in an over-arching review such as this.}
\response{While it is true that proteochemometric methodologies are
  not as well known as say docking or pharmacophore modeling, we
  choose to include them since they are an example of a (statistical)
  QSAR technique that does explicitly include ligand and receptor features}

\comment{RDKit is not mentioned at all in this article and is widely
  used as a cheminformatics toolkit in industry and academia and free
  and open-source (although from commercial beginnings).}
\response{We agree that this toolkit should have been included. We
  have updated the text on page 7 to take note of RDKit, though we
  still primarily focus on the CDK and OpenBabel. However, Table 2
  includes RDKit in the feature comparison}

\comment{The authors could more appropriately cover the graph theory
  aspects of molecular structures by correctly applying the
  terminology to which computer scientists and cheminformaticians will
  most certainly be aware}
\response{The text has been updated}

\comment{The inclusion of ontologies is not really useful here as the
  conclusions drawn are vague and do not really inform the current
  state-of-the-art and usefulness of ontologies in cheminformatics.}
\response{The text has been updated on page 4 to make the discussion
  of ontologies more focused and the use cases clearer, specifically 
  identifying challenges in cross-domain querying, classification 
  and integration within the context of whole-systems biology, and the 
  growing essental role that ontologies and semantic technologies are 
  playing in this matter.}

@@

\comment{p1, c1, l32: I think the phrase miracle molecule could be
  more usefully replaced with new drugs or new small molecule
  therapeutics.}
\response{The text has been updated to use the term ``therapeutic
  molecule''}

\comment{p2, c2, l57: it might be worthwhile here clarifying in the text
precisely the properties that require satisfaction to deliver a small
molecule therapeutic, not ?promising?. A drug must be safe and
efficacious ultimately; perhaps this should be mentioned first
followed by the typical pitfalls and how they are assessed?}
\response{The text has been updated to provide some examples of
  properties that would characterize a therapeutically useful small molecule}

\comment{P3, c1, l41: the authors need to cover the Morgan algorithm
  (published in 1965) here and explain the canonicalization
  process. The issues mentioned in different canonicalization
  implementations providing different SMILES strings is also a
  challenge with InChI codes with different softwares giving different
  representation. Therefore, this is still not a solved problem as
  suggested here. Could the authors clarify this in the text?}
\response{We have updated the text to reference the Morgan
  algorithm. Regarding the issue of InChI codes, we believe that this
  is not the case. Currently, there is only one implementation of the
  InChI algorithm and even if alternative implementations were to be
  developed, the InChI specification is publically available. As a
  result, there should be no differing InChI representations for a
  given input structure.}

\comment{P3, c2: structure-based fingerprints referred to here sound
  like Daylight-style fingerprints to this reviewer. Is this the case?
  I would emphasise the distinction of structure-key and hash-key
  fingerprints. My understanding is that Daylight-style fingerprints
  enumerate paths of 7 in length, not 8, could the authors provide a
  reference for this?}
\response{The section on fingerprints has been rewritten to be more
  clearer as well as distinguish between the two broad classes of
  fingerprints. We have also mentioned the use of fingerprints in
  similarity searches and the Tanimoto coefficient}

\comment{P3, c2, l42: is this true given the lower complexity of
  molecular graphs compared with other much larger and denser graphs?
  I think a reference to a known paper that states the problem clearly
  would be of benefit here.}
\response{While it is true that isomorphism on smaller graphs that are
  characteristic of small molecule is in an absolute sense, not very
  slow, there are a number of cases (polycyclic hydrocarbons, steroids
  etc.) that can take significantly longer. More generally, the fact
  that subgraph isomorphism is NP-complete, means that no time guarantees
  can be provided. Thus when performing this on \emph{millions} of
  molecules as in a database search, we may not be able to complete
  the operation. In practice this is usually not the case, but still,
  isomorphism algorithms do take much longer than fingerprint
  screening. Hence in practice performing isomorphism tests for large
  databases is infeasible. We have included a reference that
  explicitly talks about this problem.}

@@

\comment{P7, c2, l25: the authors should go back even further in the
  history of cheminformatics here before DENDRAL. While this is an
  excellent example of crossover, it is not by any means the founding
  of our field. Aspects of mathematical chemistry should also be
  mentioned in this article, which date back even further, such as
  1894 with the publication of ?The Principles of Mathematical
  Chemistry? published by Helm.}
\response{As noted previously, we have included a supplemental
  document that lists some of the milestones in the history of
  cheminformatics. However, we note that the suggested reference by
  Helm is not specifically related to cheminformatics. Rather, it is a
  mathematical treatment of physical chemistry. Given that DENDRAL is
  referenced in the context of structure enumeration, we feel that
  inclusion of this reference would be somewhat irrelevant. We have
  included this reference in the supplemental history}

\comment{P7, c2, l51: why is chemically in inverted commas and not
  algorithmically? I would prefer inverted commas for neither.}
\response{Quotes have been removed}

\comment{P8, c1, l10: ELN's should be ELNs.}
\response{Corrected}

\comment{P8, c1, l5: the users emphasise open source, open data but
  this should be an article on all of cheminformatics not just a
  particular subset.}
\response{We have restructured the section on
  open source and open data. It is true that we still focus on open
  source/open data, because this ensures that a reader will be able to
  explore cheminformatics problems with a minimum of hindrances}

\comment{P7, c1, l9: the authors make an interesting point here
  regarding why cheminformatics software has historically been from
  commercial suppliers. Given the scope of this manuscript it would be
  interesting to expand on this discussion.} 
\response{We have
  restructured the Conclusions section to expand on the role of
  cheminformatics in industry versus bioinformatics, and why much
  of cheminformatics tools and data have been commercial, in contrast
  to freely available tools and data in bioinformatics}

### 2011-11-18 paper.tex
Overall edit and specific changes to ontology section

Given that a fundamental principle of the field is that \emph{similar
  molecules exhibit similar properties} \cite{Johnson:1990qf}, the
choice of representation is key in determining how such similarities
are evaluated, and thus the effectiveness of subsequent analyses. But
even after one has selected a representation one faces the challenge
of balancing computational costs with the utility of the
representation. For example, a full 3D description of a molecule,
taking into account all possible conformers would let us accurately
predict many properties. But the size of this representation and the
time required to evaluate it would be prohibitive. Instead, can we
obtain accurate predictions from a subset of conformers? Or, can we
obtain comparable accuracy by using a 2D representation? And so,
what type of labels are required? Currently, many of these questions
are answered using trial and error.

@@

As noted above, in this article we choose to focus on cheminformatics
in the context of drug discovery and development. Given that this is a
slow and expensive process, the concept of ``risk minimization'' plays
an important role in this field. Thus, we structure the article around
this idea -- how to minimize the chances of a small
molecule failing, due to poor physical, chemical or biological
properties, in the various research and development stages as a drug
candidate, and how to successfully employ cheminformatics methods in
these efforts. As an example, a molecule must be soluble and show a
certain degree of bioavailability to be considered as drug
candidate. In cases where these properties are poor, a cheminformatics
approach can suggest replacement of certain functional groups, that
maintain potency but improve the solubility and bioavailability.
Table~\ref{table:properties} (in the Appendix) summarizes a number of
properties a drug candidate must satisfy to be considered
therapeutically useful (such as being appropriately soluble,
exhibiting minimal toxicity and being selective for the intended
target) and sketches the role of cheminformatics at each stage of drug
development.
%
\subsection{Representing and Searching Structures}
\label{sec:databases}
%
Most cheminformatics applications rely on large databases of chemical
structures, their properties, and relationships to, \textit{e.g.},
biological targets.  Organizing and maintaining such databases, as
well as searching and clustering similar structures together, are
essential to enable many scientific applications. Yet, each one of
those areas poses its unique computer science challenges.

There is considerable progress in open chemical and bioactivity
databases like ChEMBL and PubChem, which are both freely
available and contain in the order of millions (ChEMBL) and tens of
millions (PubChem) of data points each. Still, the integration of
chemical databases with one another remains a challenge, not only due
to the normalization problems in the chemical and bioactivity
landscape, but also due to sheer data volume. One problem is the
annotation with the ``protein targets'' a molecule binds to -- here
dozens of protein and gene identifiers exist, generally without the
availability of a one-to-one mapping from each ontology, which makes
merging databases often a cumbersome exercise.

Given the requirement to ``make sense'' of as much chemical data at
hand as possible, we have to be able to search in complex data, for
example multi-labelled molecular graphs, where graph labels can change
from database to database. Hence, we need to be able to capture
molecular graphs in a risk-bounded machine encoded way.

A fundamental challenge in cheminformatics is the unique
representation of chemical structures in a machine readable format. A
(chemical) graph can be encoded in multiple ways, depending on how one
orders the nodes, resulting in multiple representations of the same
molecule. When one considers the need to include other aspects such as
protonation states (how the molecule changes when one proton is added
to it) and tautomer states (how the molecule can change when a proton
migrates from one part of the molecule to another), the challenge
increases. The need for a unique representation that is invariant with
respect to the atom ordering arises due to the fact that graph
isomorphism (i.e., checking whether two molecules are the same) is an
expensive operation. The first such ``unique representation
algorithm,'' or \emph{canonicalization} algorithm, was described by
Morgan \cite{Morgan1965}, allowing one to generate unique string
representations of chemical graphs letting one compare structures via
string comparisons. The SMILES (Simplified Molecular Input Line
Specification) format \cite{Weininger:1988kx} is an example of a
representation that can be canonicalized. Since the original
canonicalization algorithm was proprietary, multiple implementations
of the format have become available, each one employing a different
canonicalization algorithm, usually based on the Morgan algorithm. See
Warr~\cite{Warr:2011vn} for an extensive discussion on chemical
structure representations.

With each database using their own algorithm for molecular encoding,
automated exchange of data between different databases was hindered.
As more and more data became available online, the use of unique
structure-based identifiers became a pressing need, resulting in the
IUPAC International Chemical Identifier (InChI) to be developed. The
InChI is a non-proprietary, structured textual identifier for chemical
entities~\cite{inchi}. InChI identifiers are not intended to be read
and understood by humans, but are useful for computational matching of
chemical entities. For rapid database lookups, the InChIKey is a
hashed key for the InChI, which has an invariant length of 14
characters. Figure~\ref{figure:smiles} illustrates
the SMILES, InChI and InChIKey for lipoic acid.

@@

InChI or other identity-mapping algorithms allow for exact
searching. However, two other practically relevant algorithms for
scientific discovery based on chemical databases are
\emph{substructure searching} and \emph{similarity searching}, which
are required to generalize from the data points contained in a
particular database to other, related molecules. In substructure
searching, the database is searched for a specified wholly contained
part of the structures in the database, while in similarity searching,
the database is searched for structures similar (in some structure or
property space) to a provided search, or query, structure. Chemical
search packages are often implemented and optimized for a given
database technology; for example, the OrChem package is an open source
chemical search package for the Oracle database \cite{rijnbeek2009}.

@@

\subsection{Linking Molecules to Other Data}
\label{sec:prof-ident}

The preceding section focuses on the choice of representation of
chemical structures. With the profusion of databases that contain
details of small molecules, an emerging problem is the integration of
such databases. Such integration forms the basis of ``systems'' level
investigations of biological systems. A current challenge in
cheminformatics is how to identify ``equivalent'' objects across data
sources. At one level, equivalent compounds can be identified via
canonical structure representations (InChI). But there are other
equivalences beyond structure. For example, compounds may bind to the
same or similar targets; they may be have similar biological function
(say, anti-neoplastic, that is to say, it prevents the development of
tumors) and so on. To assess such higher level characteristics,
ontologies and other semantic technologies have been employed, to first,
support annotations of chemical systems and second, enable semantic
analyses of chemical systems in conjunction with other biological
systems (such as pathways). A number of challenges remain in this area
- the use of ontologies allows one to build a semantic network of
concepts that links molecules to other molecules as well as other
objects such as targets, genes and pathways. Open questions include
the following.  How does one evaluate similarities in such a
heterogenous network \cite{couto2010}? How does one integrate chemical
structure information with ontologies \cite{hastingsowled2010}? And
more generally, how can we achieve automated inferencing in such
heterogenous linked systems, allowing us to achieve a systems level
understanding of small molecules \cite{Oprea:2007fk}.

\subsection{Expanding Chemical Space}
\label{sec:struct-enum}
Enumerating molecules is a combinatorial problem that has fascinated
chemists, computer scientists and mathematicians alike, for more than
a century. Indeed, many fundamental principles of graph theory and
combinatorics were developed in the context of counting isomers of
paraffin by Cayley, Polya and others. In the 1960's, Lederberg, 
Dejerassi and others developed algorithms to enumerate structures
based on spectral data, leading to the development of DENDRAL, widely
considered as the first expert system \cite{DENDRAL}.

@@

A key application of structure enumeration is the elucidation of
structures based on spectral data \cite{Kind:2010zr}. This is
especially relevant for identifying metabolites, small molecules
that are the byproducts of metabolic processes and which provide
insight into the biological state (diseased, fasting, etc.) of an
organism. Ideally, a chemist would gather spectral data (mass, LC/MS, etc.),
and the algorithm would provide a list of structures that can give
rise to the observed spectra. Some commercial products such as MOLGEN
(\url{http://molgen.de}) are able to perform this task in a rapid
manner.

@@

While the complexity of molecular graph enumeration remains an open
problem, an alternative to enumerating the space is to sample it
\cite{goldberg1999}. Sampling procedures based on the Metropolis or
the Genetic Algorithms have been developed and used to elucidate
natural compounds from NMR data. It should be noted that even in the
face of the computational complexity of molecular enumeration,
computational chemists have developed and successfully used
enumeration tools to generate large chemical libraries such as GDB-13,
comprising almost a billion chemical structures~\cite{GDB} (but only
considering molecules up to 13 heavy atoms and using only carbon,
nitrogen, oxygen, sulfur and chlorine). Yet, the current enumeration
software products do not generally produce stereoisomers or tautomers;
these require specific enumeration procedures, which are still the
subject of investigations by the cheminformatics community.

@@

The first step in the predictive modeling of biological activities is
to generate \emph{molecular descriptors} (a.k.a. \emph{features}) that
are numerical representations of different structural features. Here
multiple approaches exist. We can transform molecular graphs into a
numerical vector representation of chemical structures
\cite{todeschini2000}, ranging from counts of elements to eigenvalues
of the Laplacian matrix.  Alternatively, we can compare molecular
graphs directly, via \emph{kernel methods}, where a kernel on graphs
$G$ and $G'$ provides a measure of how similar $G$ is to $G'$, or a
kernel on a single graph compares measures similarities between the
graph's nodes. In these cases, rather than compute vectorial
representations, we directly operate on the graph representations.

@@

Having settled on a numerical representation and a possible class of
model types, one must address the goal of the model. Are we
looking for pure predictive power, with no interest in any explanatory
features? Or are we more interested in decent predictive power, but
also some explanation as to why a molecule is predicted as toxic or
non-toxic, for example. The former scenario is common in virtual
screening settings, where we require a high degree of accuracy and fast
predictions, but do not really care why one molecule is active
and another inactive. These types of models can be black boxes (neural
networks) and algorithmic models \cite{Breiman:2001fk} (random
forests). The latter scenario is more common in exploratory and
optimization settings, where we hope that the output of the model will
guide us in chemical modifications to improve the property being
modeled.  In this case, we need to understand the effect of a
certain structural feature on the observed potency and expect that the
model will provide insight. These types of models are generally
distributional in nature (such as linear regression and Partial Least Squares) though
some algorithmic approaches such as random forests can provide insight.

@@

One of the key challenges faced by predictive modeling is the fact
that small molecules are not static and do not exist in
isolation. Traditionally, predictive models have focused on a single
structure for a small molecule and have ignored the role of the
receptor (in receptor-mediated scenarios). Yet, small molecules can
exist in multiple tautomeric forms and usually in multiple
conformations. As a result, enhancing the accuracy of predictions
ideally will require that the 3D geometries of that molecule be taking into account and, as
far as possible, take into account the receptor
simultaneously. Multi-conformer modeling has been addressed in the
4D-QSAR methodology described by Hopfinger and co-workers
\cite{Albuquerque:1998ys}.  Recent techniques such as
\emph{multiple-instance learning} could also be applied to the
multi-conformer problem.

@@

The relevance of scientific knowledge management is increasing not
only for reducing the risk in current research, but also for enabling
new collaboration/innovation opportunities with internal partners and
a rapidly growing number of external/public partners. In
cheminformatics and chemistry, scientists already switched years
ago from paper lab-notebooks to online collaboration platforms, called
ELNs (Electronic Lab Notebooks). So, what were the drivers behind
chemists, and companies, adopting ``Enterprise 2.0,'' a social online
collaboration culture?

Before the year 2000, many chemists were still using paper
lab-notebooks, multiple compound registration, and search
tools. The overall architecture was too inflexible to
adopt to fast changing data standards, scientists spent too much time
with administrative work, chemical data quality was inconsistent, an
alignment with other working groups was inefficient (\emph{e.g.}, for
running analytical experiments). Legal disputes would require manual
searches through many paper lab notebooks and data synchronization was
painful, hampering large scale collaborations.

@@

For researchers interested in the development of new algorithms in the
field, a software library that handles chemical structures and related
data is essential. A wide variety of such chemical toolkits are
available, both commercial (which may offer free academic
licenses) and Open Source.

Due to the importance of cheminformatics to the pharmaceutical
industry in particular, numerous commercial vendors provide libraries,
applications and database cartridges to handle a wide variety of
needs including Accelrys, BioSolveIT, ChemAxon, Chemical Computing Group,
Daylight, Open Eye, Schrodinger, Tripos, and Xemistry. More recently
there has been a growth in the development of Open
Source cheminformatics software, and this presents opportunities for
the rapid development and implementation of novel algorithms which can
build on the existing Open Source ecosystem. In this article we do not debate the
arguments for and against Open Source; we focus on Open Source
tools simply because this ensures that a reader will be able to explore
cheminformatics problems with a minimum of encumbrances.

@@

    biological activity of a chemical structure in a quantitative manner.

>>>>

Given that a fundamental principle of the field is that \emph{similar
  molecules exhibit similar properties} \cite{Johnson:1990qf}, the
choice of representation is key in determining how such similarities
are evaluated, and thus the effectiveness of subsequent analyses. But
even after one has selected a representation one faces the challenge
of balancing computational costs with the utility of the
representation. For example, a full 3D description of a molecule,
taking into account all possible conformers would let us accurately
predict many properties. But the size of this representation and the
time required to evaluate it would be prohibitive. Instead, can we
obtain accurate predictions from a subset of conformers? Or, can we
obtain comparable accuracy by using a 2D representation? And if so,
what type of labels are required? Currently, many of these questions
are answered using trial and error.

@@

As noted above, in this article we choose to focus on cheminformatics
in the context of drug discovery and development. Given that this is a
slow and expensive process, the concept of ``risk minimization'' plays
an important role in this field. Thus, we structure the article around
this idea -- how to minimize the chances of a small
molecule failing, due to poor physical, chemical or biological
properties, in the various research and development stages as a drug
candidate, and how to successfully employ cheminformatics methods in
these efforts. As an example, a molecule must be soluble and show a
certain degree of bioavailability to be considered as drug
candidate. In cases where these properties are poor, a cheminformatics
approach can suggest replacement of certain functional groups, that
maintain potency but improve the solubility and bioavailability.
Table~\ref{table:properties} (in the Appendix) summarizes a number of
properties a drug candidate must satisfy to be considered
therapeutically useful (such as being appropriately soluble,
exhibiting minimal toxicity and being selective for the intended
target) and sketches the role of cheminformatics at each stage of drug
development.
%
\subsection{Representing and Searching Structures}
\label{sec:databases}
%
Most cheminformatics applications rely on large databases of chemical
structures, their properties, and relationships to, \textit{e.g.},
biological targets.  Organizing and maintaining such databases, as
well as searching and clustering similar structures together, are
essential to enable many scientific applications. Yet, each one of
these areas poses its unique computer science challenges.

There is considerable progress in open chemical and bioactivity
databases like ChEMBL and PubChem, which are both freely
available and contain in the order of millions (ChEMBL) and tens of
millions (PubChem) of data points each. Integration of these disparate
data is essential for researchers to obtain the fullest possible 
perspective on what is presently known, tracking ongoing advances in 
science as these become available. Still, the integration of
chemical databases with one another remains a challenge, not only due
to the normalization problems in the chemical and bioactivity
landscape, but also due to sheer data volume. One problem is the
annotation of the ``protein targets'' a molecule binds to, as  
dozens of protein and gene identifiers exist, generally without the
availability of a one-to-one mapping from each ontology, which makes
merging databases often a cumbersome exercise.
% [JH comment] 
% Is this above sentence really true? I thought there was general standardization on 
% UniProt for proteins, GenBank for genes, and that mappings between the other
% sorts of identifiers to these primary ones were actually well maintained?
% As I understand it, the main problem with target representation is that
% protein/gene targets are often unknown, and then it is necessary to map to
% pathway, cell, tissue, biological process etc. --> representing the maximum
% that is known about the mechanism of action for the molecule -- and THESE
% data types are often not well standardized
% [JH comment]

Given the requirement to ``make sense'' of as much chemical data at
hand as possible, we have to be able to search in complex data, for
example multi-labelled molecular graphs, where graph labels can change
from database to database. Hence, we need to be able to capture
molecular graphs in a risk-bounded machine encoded way. %risk-bounded meaning here not clear

A fundamental challenge in cheminformatics is the unique
representation of chemical structures in a machine readable format. A
(chemical) graph can be encoded in multiple ways, depending on how one
orders the nodes, resulting in multiple representations of the same
molecule. When one considers the need to include other aspects such as
protonation states (how the molecule changes when one proton is added
to it) and tautomer states (how the molecule can change when a proton
migrates from one part of the molecule to another), the challenge  %This is repeating something stated earlier
increases. The need for a unique representation that is invariant with
respect to the atom ordering arises due to the fact that graph
isomorphism (i.e., checking whether two structures represent the same molecule) is an
expensive operation. The first such ``unique representation
algorithm,'' or \emph{canonicalization} algorithm, was described by
Morgan \cite{Morgan1965}, allowing one to generate unique string
representations of chemical graphs letting one compare structures via
string comparisons. The SMILES (Simplified Molecular Input Line
Specification) format \cite{Weininger:1988kx} is an example of a
representation that can be canonicalized. Since the original
canonicalization algorithm was proprietary, multiple implementations
of the format have become available, each one employing a different
canonicalization algorithm, usually based on the Morgan algorithm. See
Warr~\cite{Warr:2011vn} for an extensive discussion on chemical
structure representations.

With each database using their own algorithm for molecular encoding,
automated exchange of data between different databases was hindered.
As more and more data became available online, the use of unique
structure-based identifiers became a pressing need, resulting in the
development of the IUPAC International Chemical Identifier (InChI). The
InChI is a non-proprietary, structured textual identifier for chemical
entities~\cite{inchi}. InChI identifiers are not intended to be read
and understood by humans, but are useful for computational matching of
chemical entities. For rapid database lookups, the InChIKey is a
hashed key for the InChI, which has an invariant length of 14
characters. Figure~\ref{figure:smiles} illustrates
the SMILES, InChI and InChIKey for lipoic acid.

@@

InChI or other identity-mapping algorithms allow for exact
searching. However, two other practically relevant algorithms for
scientific discovery based on chemical databases are
\emph{substructure searching} and \emph{similarity searching}, which
are required to generalize from the data points contained in a
particular database to other, related molecules. In substructure
searching, the database is searched for a specified wholly contained
part of the structures in the database, while in similarity searching,
the database is searched for structures similar (in some structure or
property space) to a provided search, or query, structure. Chemical
search packages are often implemented and optimized for a given
database technology, for example, the OrChem package is an open source
chemical search package for the Oracle database \cite{rijnbeek2009}.

@@

\subsection{Situating Molecular Data in Biological Context}
\label{sec:prof-ident}

The preceding section focuses on the choice of representation of
chemical structures. In the quest for the discovery of novel
therapeutic agents, it is essential to study molecules as they are
situated in a biological context, underlying the processes of life, 
as metabolites, cofactors for enzymatic reactions, hormones, signalling
molecules and environmental xenobiotics. This gives rise to the need
to adequately represent and organise chemical data across biological
databases such as those containing pathways, protein information or 
biological samples. Integration and processing of data from such 
disparate domains underlies ``systems'' level biological research.
This gives rise to some additional challenges that standard chemical
structural representation cannot address: firstly, the representation of
\textit{classes of chemical entities} -- since, in many cases, a class of 
compounds behave in a certain context in a biological system rather than
a single molecule with a fully-specified structure -- and for similar
reasons, chemical concepts such as \textit{groups} (parts of molecules); 
secondly, the need to represent \textit{non-structural} groupings of 
compounds of interest, since compounds may bind to the same or similar 
targets or they may have  similar biological functions such as acting as 
an anti-neoplastic -- preventing the development of tumors. With the 
many different biological databases needing to refer to these different
chemical entities in an organised and standardized 
fashion to enable cross-database integration for a whole-systems 
perspective, \textit{ontologies} and other semantic technologies 
are being successfully employed to support annotation, classification and
semantic cross-domain querying of chemical entities within
biological systems. The most widely used ontology for biologically relevant
chemical entities is ChEBI \cite{chebi2010}, containing around 26,500 
entities in an interlinked semantic graph as of November 2011, and
being actively used for annotation in dozens of biological databases. 
Ontologies are based on logical languages of 
varying levels of expressivity, accompanied by sophisticated reasoning
algorithms. They can be used to construct various sorts of semantic
similarity (e.g. similarity in function, similarity in application) 
which complement traditional structural similarity methods \cite{couto2010}.
An open challenge is integrating graph-based chemical
structure information with the wider logic-based ontology information, which 
will allow combined automated reasoning and querying over the different domains 
encoded in a growing number of interlinked ontologies \cite{hastingsowled2010}. 
Enabling automated inferencing in such heterogenous linked systems brings us closer to
achieving a systems level understanding of bioactive chemistry \cite{Oprea:2007fk}.

\subsection{Expanding Chemical Space}
\label{sec:struct-enum}
Enumerating molecules is a combinatorial problem that has fascinated
chemists, computer scientists and mathematicians alike, for more than
a century. Indeed, many fundamental principles of graph theory and
combinatorics were developed in the context of counting isomers of
paraffin by Cayley, Polya and others. In the 1960's, Lederberg, 
Djerassi and others developed algorithms to enumerate structures
based on spectral data, leading to the development of DENDRAL, widely
considered as the first expert system \cite{DENDRAL}.

@@

A key application of structure enumeration is the elucidation of
structures based on spectral data \cite{Kind:2010zr}. This is
especially relevant for identifying metabolites, small molecules
that are the byproducts of metabolic processes and which provide
insight into the biological state (diseased, fasting, etc.) of an
organism. Ideally, a chemist would gather spectral data (mass, LC/MS, etc.),
and the algorithm would provide a list of structures that can give
rise to the observed spectra. Some commercial products such as MOLGEN
(\url{http://molgen.de}) are able to perform this task rapidly.

@@

While the complexity of molecular graph enumeration remains an open
problem, 
% [JH comment]
% Not sure what is meant here? Surely, the complexity of molecular graph enumeration 
% is upper-bounded exponentially in the number of atoms? Additional complexity
% arises from algorithms to select "good" candidates from the candidates?
% Perhaps what is meant is not that the complexity itself is an open problem,
% but finding methods for enumeration which are more efficient is an open problem?
% Alternatively, if complexity in the CS sense is not what is meant here, maybe it will
% help this audience if a different word is used. 
% [JH comment]
an alternative to enumerating the space is to sample it
\cite{goldberg1999}. Sampling procedures based on the Metropolis or
Genetic Algorithms have been developed and used to elucidate
compounds from NMR data. It should be noted that even in the
face of the computational complexity of molecular enumeration,
computational chemists have developed and successfully used
enumeration tools to generate large chemical libraries such as GDB-13,
comprising almost a billion chemical structures~\cite{GDB} (but only
considering molecules up to 13 heavy atoms and using only carbon,
nitrogen, oxygen, sulfur and chlorine). Yet, the current enumeration
software products do not generally produce stereoisomers or tautomers;
these require specific enumeration procedures, which are still the
subject of research by the cheminformatics community.

@@

The first step in the predictive modeling of biological activities is
to generate \emph{molecular descriptors} (a.k.a. \emph{features}) that
are numerical representations of different structural features. Here
multiple approaches exist. We can transform molecular graphs into a
numerical vector representation of chemical structures
\cite{todeschini2000}, ranging from counts of elements to eigenvalues
of the Laplacian matrix.  Alternatively, we can compare molecular
graphs directly, via \emph{kernel methods}, where a kernel on graphs
$G$ and $G'$ provides a measure of how similar $G$ is to $G'$, or a
kernel on a single graph compares measured similarities between the
graph's nodes. In these cases, rather than compute vectorial
representations, we directly operate on the graph representations.

@@

Having settled on a numerical representation and a possible class of
model types, one must address the goal of the model. Are we
looking for pure predictive power, with no interest in any explanatory
features? Or are we more interested in decent predictive power, but
also some explanation as to why a molecule is predicted as toxic or
non-toxic (for example)? The former scenario is common in virtual
screening settings, where we require a high degree of accuracy and fast
predictions, but do not really care why one molecule is active
and another inactive. These types of models can be black boxes (neural
networks) and algorithmic models \cite{Breiman:2001fk} (random
forests). The latter scenario is more common in exploratory and
optimization settings, where we hope that the output of the model will
guide us in chemical modifications to improve the property being
modeled.  In this case, we need to understand the effect of a
certain structural feature on the observed potency and expect that the
model will provide insight. These types of models are generally
distributional in nature (such as linear regression and Partial Least Squares) though
some algorithmic approaches such as random forests can provide insight.

@@

One of the key challenges faced by predictive modeling is the fact
that small molecules are not static and do not exist in isolation. 
Traditionally, predictive models have focused on a single
structure for a small molecule and have ignored the role of the
receptor (in receptor-mediated scenarios). Yet, small molecules can
exist in multiple tautomeric forms and usually in multiple
conformations. As a result, enhancing the accuracy of predictions
ideally will require that the 3D geometries of that molecule be taken 
into account and, as far as possible, that the receptor also be taken
into account simultaneously. Multi-conformer modeling has been addressed in the
4D-QSAR methodology described by Hopfinger and co-workers
\cite{Albuquerque:1998ys}.  Recent techniques such as
\emph{multiple-instance learning} could also be applied to the
multi-conformer problem.

@@

The relevance of scientific knowledge management is increasing not
only for reducing the risk in current research, but also for enabling
new collaboration and innovation opportunities with internal partners and
a rapidly growing number of external/public partners. In
cheminformatics and chemistry, scientists already switched years
ago from paper lab-notebooks to online collaboration platforms, called
ELNs (Electronic Lab Notebooks). So, what were the drivers behind
chemists, and companies, adopting ``Enterprise 2.0,'' a social online
collaboration culture?

Before the year 2000, many chemists were still using paper
lab-notebooks, multiple compound registration, and search
tools. The overall architecture was too inflexible to
adapt to fast-changing data standards, scientists spent too much time
with administrative work, chemical data quality was inconsistent, and
alignment with other working groups was inefficient (\emph{e.g.}, for
running analytical experiments). Legal disputes would require manual
searches through many paper lab notebooks and data synchronization was
painful, hampering large scale collaborations.

@@

For researchers interested in the development of new algorithms in the
field, a software library that handles chemical structures and related
data is essential. A wide variety of such chemical toolkits are
available, both commercial (which may offer free academic
licenses) and Open Source.
% [JH comment]
% we should standardize our nomenclature for the phrase 'Open Source' in the
% document. Here it is capitalized, earlier it has been uncapitalized 
% ('open source'), and below it is hyphenated! (open-source)
% [JH comment]

Due to the importance of cheminformatics to the pharmaceutical
industry in particular, numerous commercial vendors provide libraries,
applications and database cartridges to handle a wide variety of
needs including Accelrys, BioSolveIT, ChemAxon, Chemical Computing Group,
Daylight, Open Eye, Schrodinger, Tripos, and Xemistry. More recently
there has been a growth in the development of Open
Source cheminformatics software, and this presents opportunities for
the rapid development and implementation of novel algorithms which can
build on the existing Open Source ecosystem. In this article we do not debate the
arguments for and against Open Source; we focus on Open Source
tools simply because this ensures that any interested reader will be able to explore
cheminformatics problems with a minimum of hindrances.

@@

    biological activity of a chemical structure in a quantitative fashion.
